\chapter*{Acknowledgements}
In writing this thesis, I have received help from a great many people. First and foremost I would like to thank Marko van Eekelen, Hendrik Tews and Sjaak Smetsers for the great assistance they have given me. They have helped me with many different aspects of this thesis, from the verification of individual lemmas to planning the research itself. Secondly, I want to thank Ren{\'e} Reusner and Adam Lackorzynski for never becoming tired of answering all my questions regarding the Fiasco IPC implementation. In this regard, the L4 hackers mailinglist has also been very helpful; their responses were always fast and to the point. The structure and contents of this document have been extensively commented upon by Jens Wagemakers, for which I owe him many thanks. For a more general discussion of my thesis I often turned to Sietse Overbeek, with whom I had some very useful discussions. When a technical problem arose, Engelbert Hubbers was always quick to solve it.\emptyline

Besides the abovementioned people, who were directly involved with the thesis' contents, I have also received support from a lot of people that did not involve the contents of this thesis. Most notably, these include my friends, colleagues and family and of course my girlfriend, Jeltine Jungheim. I want to specifically mention my dad who has always supported me in whatever I did, but unfortunately passed away far too young; this thesis is dedicated to him.\emptyline

I hereby apologize to anyone that has not made it in this list, that does not mean I did not appreciate your help but merely that my memory has failed me.